---
title: "01-basics"
format: html
---

# Introduction

R is a programming language designed primarily for statistical computing, data analysis, and visualization.

This chapter introduces the basic syntax and operations used in R.

# Variables and Assignment

R commonly uses `<-` operator for assignment.

```{r}
x <- 10
name <- "Sujan"
is_human <- TRUE
salary <- 0.1
```

We can then inspect these values:

```{r}
x
name
is_human
salary
```

## Assignment Operator

We can also use `=` operator for assignment:

```{r}
x = 20
x
```

However, `<-` is the most conventional assignment operator in R.

# Arithmetic Operators

```{r}
10 + 5
10 - 5
10 * 5
10 / 5
10^2
10 %% 3
10 %/% 3
```

| Operator | Meaning            |
|----------|--------------------|
| `+`      | Addition           |
| `-`      | Subtraction        |
| `*`      | Multiplication     |
| `/`      | Division           |
| `^`      | Exponentiation     |
| `%%`     | Modulo / remainder |
| `%/%`    | Integer division   |

# Comments

Comments begin with `#`.

```{r}
# This is a comment

x <- 10 # This assigns 10 to x
```

# Getting Help

R has extensive built-in documentation.

```         
?mean
```

You can also use:

```         
help(mean)
```

To search for a topic:

```         
??regression
```

# Basic Output

Use `print()` to explicitly print a value.

```{r}
print("Hello R!")
```

However, R can also display an expression directly:

```{r}
x <- 30
x
```

# Checking the working Directory

```{r}
getwd()
```

Change it when necessary:

```         
setwd("path/to/directory")
```
> **Recommendation:** When working with RStudio Projects, prefer project-relative paths instead of repeatedly using `setwd()`.

# Quick Reference

| Command | Purpose |
|---|---|
| `<-` | Assignment |
| `print()` | Print a value |
| `getwd()` | Get working directory |
| `setwd()` | Change working directory |
| `?function` | Open function documentation |
| `help()` | Open documentation |
| `??term` | Search documentation |
| `#` | Comment |