---
title: "Data Types"
format: html
---

# Introduction

Every value in R has a type. Understanding data types is important because
different types behave differently when we perform calculations, comparisons,
or transformations.

The most common basic data types in R are:

- Numeric
- Integer
- Character
- Logical
- Complex

R also has special values such as `NA`, `NULL`, `NaN`, and `Inf`.

# Numeric

Numeric values represent numbers with decimal values.
```{r}
x <- 10
y <- 3.14

x
y
```

Check the type:

```{r}
typeof(x)
typeof(y)
```

Both are stored as:

```text
double
```

In R, regular numeric values are generally stored as double-precision
floating-point numbers.

## Example

```{r}
price <- 19.99
temperature <- 23.5
average <- 87.25
```

# Integer

Integers represent whole numbers.

An integer can be explicitly created by adding `L`:

```{r}
x <- 10L

x
typeof(x)
```

The result is:

```text
integer
```

Compare:

```{r}
x <- 10
y <- 10L

typeof(x)
typeof(y)
```

Output:

```text
double
integer
```

> **Remember:** `10` is normally stored as a `double`, while `10L` is stored
> as an `integer`.

# Character

Character values represent text.

They can be created using either single or double quotation marks.

```{r}
name <- "Sujan"
country <- 'Germany'

name
country
```

Check the type:

```{r}
typeof(name)
```

Output:

```text
character
```

## Combining strings

Use `paste()`:

```{r}
first_name <- "Sujan"
last_name <- "Karna"

paste(first_name, last_name)
```

Or use `paste0()`:

```{r}
paste0(first_name, "_", last_name)
```

The difference is that `paste()` inserts a separator by default,
while `paste0()` does not.

```{r}
paste("Hello", "World")
paste0("Hello", "World")
```

# Logical

Logical values represent Boolean values.

R has two logical values:

```r
TRUE
FALSE
```

Example:

```{r}
is_student <- TRUE
has_license <- FALSE

is_student
has_license
```

Check the type:

```{r}
typeof(is_student)
```

Output:

```text
logical
```

## Logical expressions

Comparisons produce logical values:

```{r}
x <- 20

x > 10
x < 10
x == 20
x != 20
```

Output:

```text
TRUE
FALSE
TRUE
FALSE
```

# Complex

R supports complex numbers.

```{r}
z <- 2 + 3i

z
typeof(z)
```

The `i` represents the imaginary component.

You can access the real and imaginary parts:

```{r}
Re(z)
Im(z)
```

# Checking Data Types

R provides several functions for inspecting objects.

## `typeof()`

`typeof()` tells you the underlying storage type.

```{r}
x <- 10
typeof(x)
```

## `class()`

`class()` tells you the object's class.

```{r}
x <- 10
class(x)
```

For many basic objects, `typeof()` and `class()` may give similar results,
but they answer different questions.

## Type-checking functions

R provides functions that return `TRUE` or `FALSE`.

```{r}
is.numeric(10)
is.character("Hello")
is.logical(TRUE)
is.integer(10L)
```

Some other useful functions:

```{r}
is.double(10)
is.complex(2 + 3i)
```

# Type Conversion

R allows values to be converted from one type to another.

## Convert to numeric

```{r}
x <- "100"

as.numeric(x)
```

Check:

```{r}
typeof(as.numeric(x))
```

## Convert to character

```{r}
x <- 100

as.character(x)
```

## Convert to integer

```{r}
x <- 10.5

as.integer(x)
```

Note that conversion to integer removes the decimal part rather than
rounding the value.

```{r}
as.integer(10.9)
as.integer(10.1)
```

## Convert to logical

```{r}
as.logical(1)
as.logical(0)
```

# Type Coercion

R automatically converts values when different types are combined.

Consider:

```{r}
x <- c(1, 2, 3)

typeof(x)
```

The vector is numeric.

Now combine a number with a character:

```{r}
x <- c(1, 2, "3")

x
typeof(x)
```

R converts the numbers to character values.

This is called **type coercion**.

A simplified coercion hierarchy is:

```text
logical → numeric → character
```

For example:

```{r}
x <- c(TRUE, 10)

x
typeof(x)
```

The logical value is converted to numeric.

# Special Values

R has several special values that are important when working with data.

## `NA`

`NA` represents a missing value.

```{r}
x <- c(10, 20, NA, 40)

x
```

Check for missing values:

```{r}
is.na(x)
```

This returns:

```text
FALSE FALSE TRUE FALSE
```

## `NULL`

`NULL` represents the absence of a value.

```{r}
x <- NULL

x
```

Check:

```{r}
is.null(x)
```

`NULL` is different from `NA`.

- `NA` means a value is missing.
- `NULL` means there is no value.

## `NaN`

`NaN` means **Not a Number**.

It can occur when an undefined numerical operation is performed.

```{r}
0 / 0
```

Check:

```{r}
is.nan(0 / 0)
```

## `Inf`

`Inf` represents positive infinity.

```{r}
1 / 0
```

Negative infinity:

```{r}
-1 / 0
```

Check:

```{r}
is.infinite(1 / 0)
```

# Comparing Special Values

A common mistake is trying to compare `NA` using `==`.

For example:

```{r}
x <- NA

x == NA
```

The result is:

```text
NA
```

Instead, use:

```{r}
is.na(x)
```

Similarly:

```{r}
is.nan(NaN)
is.infinite(Inf)
is.null(NULL)
```

# Summary of Basic Data Types

| Type | Example | `typeof()` |
|---|---|---|
| Numeric | `10.5` | `double` |
| Integer | `10L` | `integer` |
| Character | `"Hello"` | `character` |
| Logical | `TRUE` | `logical` |
| Complex | `2 + 3i` | `complex` |

# Important Special Values

| Value | Meaning |
|---|---|
| `NA` | Missing value |
| `NULL` | No value |
| `NaN` | Not a Number |
| `Inf` | Positive infinity |
| `-Inf` | Negative infinity |

# Useful Commands

| Command | Purpose |
|---|---|
| `typeof()` | Get underlying data type |
| `class()` | Get object's class |
| `is.numeric()` | Check if numeric |
| `is.character()` | Check if character |
| `is.logical()` | Check if logical |
| `is.integer()` | Check if integer |
| `is.double()` | Check if double |
| `is.complex()` | Check if complex |
| `is.na()` | Check for missing values |
| `is.null()` | Check for `NULL` |
| `is.nan()` | Check for `NaN` |
| `is.infinite()` | Check for infinity |
| `as.numeric()` | Convert to numeric |
| `as.character()` | Convert to character |
| `as.integer()` | Convert to integer |
| `as.logical()` | Convert to logical |

# Quick Practice

Try to predict the result before running each command.

## Exercise 1

```{r}
x <- 25
typeof(x)
```

## Exercise 2

```{r}
x <- 25L
typeof(x)
```

## Exercise 3

```{r}
x <- c(1, 2, "3")
typeof(x)
```

## Exercise 4

```{r}
x <- c(TRUE, FALSE, TRUE)
typeof(x)
```

## Exercise 5

```{r}
x <- c(10, NA, 30)

is.na(x)
```

# Key Takeaways

- R has several basic data types, including numeric, integer, character,
  logical, and complex.
- `typeof()` identifies the underlying storage type.
- `class()` identifies an object's class.
- `is.*()` functions are useful for checking types and special values.
- `as.*()` functions are used for type conversion.
- R automatically performs type coercion when combining different types.
- `NA` represents missing data and should be checked using `is.na()`.
- `NULL`, `NA`, `NaN`, and `Inf` have different meanings.