---
title: "05-lists"
format: html
---

# Introduction

A **list** is a flexible R data structure that can contain multiple objects,
including objects of different types.

Unlike a vector, which normally contains values of the same basic type, a
list can contain:

- Numbers
- Character strings
- Logical values
- Vectors
- Matrices
- Other lists
- Data frames
- Functions
- Other R objects

For example:

```{r}
person <- list(
  name = "Sujan",
  age = 25,
  student = TRUE
)

person
```

This list contains a character value, a numeric value, and a logical value.

> **Key idea:** A vector is usually homogeneous, while a list can contain
> heterogeneous objects.

# Creating a List

Use the `list()` function.

```{r}
my_list <- list(
  10,
  "Hello",
  TRUE
)

my_list
```

The three elements have different types.

Check the type:

```{r}
typeof(my_list)
```

The result is:

```text
list
```

# Named Lists

It is usually more useful to give list elements names.

```{r}
person <- list(
  name = "Sujan",
  age = 25,
  student = TRUE
)

person
```

The names can be retrieved using:

```{r}
names(person)
```

# Accessing List Elements

There are three important ways to work with list elements:

```text
$
[ ]
[[ ]]
```

Understanding the difference between them is essential.

# Using `[[ ]]`

`[[ ]]` extracts the **actual element** from a list.

```{r}
person <- list(
  name = "Sujan",
  age = 25,
  student = TRUE
)

person[[1]]
```

Access by name:

```{r}
person[["name"]]
```

Access age:

```{r}
person[["age"]]
```

The returned object is the actual value.

For example:

```{r}
typeof(person[["age"]])
```

# Using `$`

The `$` operator is commonly used with named lists.

```{r}
person$name
person$age
person$student
```

This is equivalent to:

```{r}
person[["name"]]
person[["age"]]
person[["student"]]
```

For named elements, these are usually interchangeable:

```r
person$name
person[["name"]]
```

> **Tip:** `$` is convenient and readable, while `[[ ]]` is more flexible
> when the element name is stored in another variable.

# Using `[ ]`

Single brackets select a **sublist**.

```{r}
person[1]
```

Compare:

```{r}
person[1]
person[[1]]
```

The first returns a list containing the first element.

The second returns the actual value.

This distinction is extremely important.

## Example

```{r}
person[1]
```

returns a list.

```{r}
person[[1]]
```

returns:

```text
"Sujan"
```

# `[ ]` vs `[[ ]]` vs `$`

Consider:

```{r}
person <- list(
  name = "Sujan",
  age = 25
)
```

| Syntax | Result |
|---|---|
| `person[1]` | A list containing the first element |
| `person[[1]]` | The first element |
| `person["name"]` | A list containing `name` |
| `person[["name"]]` | The value of `name` |
| `person$name` | The value of `name` |

A useful way to remember it:

```text
[ ]       → subset of a list
[[ ]]     → extract an element
$         → extract a named element
```

# Lists Can Contain Vectors

A list can contain vectors.

```{r}
student <- list(
  name = "Sujan",
  scores = c(80, 90, 85)
)

student
```

Access the vector:

```{r}
student$scores
```

Then access an individual score:

```{r}
student$scores[1]
```

This combines two indexing operations:

```text
student
   ↓
$scores
   ↓
[1]
```

# Lists Can Contain Matrices

A list can contain a matrix.

```{r}
data <- list(
  name = "Scores",
  matrix = matrix(
    1:6,
    nrow = 2,
    byrow = TRUE
  )
)

data
```

Access the matrix:

```{r}
data$matrix
```

Access an element of the matrix:

```{r}
data$matrix[1, 2]
```

# Lists Can Contain Data Frames

Lists can also contain data frames.

```{r}
student_data <- data.frame(
  name = c("Alice", "Bob"),
  score = c(85, 90)
)

student <- list(
  name = "Student Information",
  data = student_data
)

student$data
```

# Nested Lists

Lists can contain other lists.

```{r}
person <- list(
  name = "Sujan",
  contact = list(
    email = "example@email.com",
    phone = "123456789"
  )
)

person
```

Access the nested list:

```{r}
person$contact
```

Access the email:

```{r}
person$contact$email
```

Or:

```{r}
person[["contact"]][["email"]]
```

Nested structures are common in real R objects.

# Length of a List

Use `length()`:

```{r}
person <- list(
  name = "Sujan",
  age = 25,
  student = TRUE
)

length(person)
```

This returns the number of top-level elements.

# Structure of a List

Use `str()` to inspect the structure.

```{r}
str(person)
```

`str()` is one of the most useful functions when working with complex R
objects.

For example:

```{r}
student <- list(
  name = "Sujan",
  age = 25,
  scores = c(80, 90, 85),
  passed = TRUE
)

str(student)
```

It shows the names, types, and structure of the elements.

> **Tip:** When you receive an unfamiliar R object, try `str(object)` first.

# Modifying List Elements

You can modify an existing element.

```{r}
person <- list(
  name = "Sujan",
  age = 25
)

person$age <- 26

person
```

Modify using `[[ ]]`:

```{r}
person[["age"]] <- 27

person
```

# Adding New Elements

You can add an element using `$`.

```{r}
person$country <- "Germany"

person
```

Or:

```{r}
person[["university"]] <- "TU Chemnitz"

person
```

# Removing Elements

Assign `NULL` to remove a named list element.

```{r}
person <- list(
  name = "Sujan",
  age = 25,
  country = "Germany"
)

person$country <- NULL

person
```

Check the result:

```{r}
names(person)
```

# Combining Lists

Use `c()` to combine lists.

```{r}
list1 <- list(
  name = "Alice",
  age = 23
)

list2 <- list(
  city = "Chemnitz",
  student = TRUE
)

combined <- c(list1, list2)

combined
```

# Repeating Lists

Use `rep()` to repeat list elements.

```{r}
x <- list(
  name = "Sujan",
  age = 25
)

rep(x, times = 2)
```

# Converting a List to a Vector

If all elements are compatible, a list can sometimes be simplified into a
vector.

```{r}
x <- list(10, 20, 30)

unlist(x)
```

Result:

```text
10 20 30
```

For a character list:

```{r}
x <- list(
  "Alice",
  "Bob",
  "Charlie"
)

unlist(x)
```

> **Important:** `unlist()` removes the list structure and attempts to create
> a simpler vector.

# Named Lists and `unlist()`

Consider:

```{r}
person <- list(
  name = "Sujan",
  age = 25
)

unlist(person)
```

The result is a vector.

Because the vector must have a single basic type, R may perform type
coercion.

# Checking Whether an Object Is a List

Use:

```{r}
is.list(person)
```

You can also use:

```{r}
typeof(person)
class(person)
```

# Lists and Vectors

A vector normally contains values of one basic type:

```{r}
numbers <- c(10, 20, 30)

typeof(numbers)
```

A list can contain different types:

```{r}
mixed <- list(
  10,
  "Hello",
  TRUE,
  c(1, 2, 3)
)

mixed
```

Compare:

```text
Vector:
[10, 20, 30]

List:
[10, "Hello", TRUE, c(1, 2, 3)]
```

# Lists and Matrices

A matrix has a fixed two-dimensional structure.

```{r}
m <- matrix(
  1:6,
  nrow = 2,
  byrow = TRUE
)

m
```

A list can contain multiple objects of completely different structures.

```{r}
objects <- list(
  numbers = c(1, 2, 3),
  matrix = m,
  text = "Hello",
  flag = TRUE
)

objects
```

# Lists and Data Frames

A data frame is closely related to a list.

Each column of a data frame can have a different type.

For example:

```{r}
df <- data.frame(
  name = c("Alice", "Bob"),
  age = c(23, 25),
  passed = c(TRUE, TRUE)
)

df
```

Conceptually, a data frame can be thought of as a collection of columns,
where each column has the same number of rows.

You can inspect its structure:

```{r}
str(df)
```

This relationship between lists and data frames will become important in the
next chapter.

# List Indexing — Important Examples

Consider:

```{r}
student <- list(
  name = "Sujan",
  scores = c(80, 90, 85),
  passed = TRUE
)
```

## First element

```{r}
student[1]
```

Returns a list.

## First element's value

```{r}
student[[1]]
```

Returns the actual value.

## Name

```{r}
student$name
```

## Name using `[[ ]]`

```{r}
student[["name"]]
```

## First score

```{r}
student$scores[1]
```

## All scores

```{r}
student$scores
```

## Second score

```{r}
student[["scores"]][2]
```

# Lists with Variables as Names

Suppose the element name is stored in a variable.

```{r}
student <- list(
  name = "Sujan",
  age = 25
)

field <- "name"
```

Using `$`:

```r
student$field
```

This looks for an element literally named `field`.

Instead, use `[[ ]]`:

```{r}
student[[field]]
```

This retrieves the element whose name is stored in `field`.

> **Important:** `[[ ]]` is useful when the name or position is stored in a
> variable.

# Useful List Functions

## `names()`

Get or set names.

```{r}
student <- list(
  "Sujan",
  25,
  TRUE
)

names(student) <- c(
  "name",
  "age",
  "student"
)

student
```

## `length()`

Get the number of top-level elements.

```{r}
length(student)
```

## `str()`

Inspect the structure.

```{r}
str(student)
```

## `unlist()`

Convert a list into a simpler vector when possible.

```{r}
unlist(student)
```

## `is.list()`

Check whether an object is a list.

```{r}
is.list(student)
```

# Applying Functions to Lists

R provides several functions for applying operations to list elements.

One of the most useful is `lapply()`.

```{r}
numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(7, 8, 9)
)

lapply(numbers, mean)
```

This calculates the mean of each vector.

The result is another list.

# `sapply()`

`sapply()` is similar to `lapply()` but tries to simplify the result.

```{r}
sapply(numbers, mean)
```

The result may be a vector instead of a list.

Compare:

```{r}
lapply(numbers, mean)
sapply(numbers, mean)
```

> **For beginners:** Learn `lapply()` first. Later, `sapply()` and other
> `apply`-family functions can be explored in more detail.

# List Function Reference

| Function | Purpose |
|---|---|
| `list()` | Create a list |
| `length()` | Number of top-level elements |
| `names()` | Get or set element names |
| `str()` | Inspect structure |
| `is.list()` | Check whether object is a list |
| `unlist()` | Simplify list into vector |
| `c()` | Combine lists |
| `rep()` | Repeat a list |
| `lapply()` | Apply function and return list |
| `sapply()` | Apply function and simplify result |

# List Indexing Reference

Given:

```{r}
student <- list(
  name = "Sujan",
  age = 25,
  scores = c(80, 90, 85)
)
```

| Operation | Code | Result |
|---|---|---|
| First element as list | `student[1]` | List |
| First element | `student[[1]]` | Value |
| Name | `student$name` | Value |
| Name with `[[ ]]` | `student[["name"]]` | Value |
| Scores | `student$scores` | Vector |
| First score | `student$scores[1]` | Value |
| Scores with `[[ ]]` | `student[["scores"]]` | Vector |
| Second score | `student[["scores"]][2]` | Value |

# Python Comparison

If you know Python, a basic R list is somewhat similar to a Python list:

| Operation | R | Python |
|---|---|---|
| Create | `list(1, "hello", TRUE)` | `[1, "hello", True]` |
| First element | `x[[1]]` | `x[0]` |
| Length | `length(x)` | `len(x)` |
| Add element | `x[[4]] <- value` | `x.append(value)` |
| Remove | `x[[2]] <- NULL` | `del x[1]` |

However, R's `$` and named-list behavior makes R lists particularly useful
for structured objects.

# Practice Exercises

## Exercise 1 — Create a List

Create a list called `student` containing:

- Name
- Age
- University
- A vector containing three exam scores
- A logical value indicating whether the student passed

Then:

1. Print the list.
2. Print its structure using `str()`.
3. Print the student's name.
4. Print the second exam score.
5. Print the number of elements.

## Exercise 2 — Modify a List

Using your `student` list:

1. Change the age.
2. Add a country.
3. Add another exam score.
4. Remove the country.

## Exercise 3 — Nested List

Create:

```text
student
├── name
├── contact
│   ├── email
│   └── phone
└── scores
```

Then retrieve:

1. Student name
2. Email
3. Phone
4. First score

## Exercise 4 — `lapply()`

Create:

```{r}
numbers <- list(
  c(10, 20, 30),
  c(5, 15, 25),
  c(100, 200, 300)
)
```

Use `lapply()` to calculate:

1. Mean
2. Maximum
3. Minimum

Then use `sapply()` and compare the results.

# Key Takeaways

- A list can contain objects of different types.
- Lists can contain vectors, matrices, data frames, and other lists.
- `list()` creates a list.
- `$` accesses a named element.
- `[[ ]]` extracts an individual element.
- `[ ]` returns a sublist.
- Understanding `[ ]`, `[[ ]]`, and `$` is essential when working with lists.
- Lists can be nested.
- `str()` is extremely useful for understanding complex objects.
- `unlist()` can simplify a list into a vector.
- `lapply()` applies a function to each list element and returns a list.
- Data frames are closely related to lists, which will become clearer in the
  next chapter.